\documentclass[a4paper]{scrreprt}

\usepackage[T1]{fontenc}
\usepackage{ae,aecompl,aeguill} 
\usepackage[latin1]{inputenc}
\usepackage{listings}
\usepackage[american]{babel}

\usepackage{todonotes}




%%\usepackage{helvet}


\lstdefinelanguage{Java}
  {morekeywords={abstract, continue, for, new, switch, assert, default, goto, package, synchronized, boolean, do, if, private, this, break, double, implements, protected, throw, byte, else, import, public, throws, case, enum, instanceof, return, transient, catch, extends, int, short, try, char, final, interface, static, void, class, finally, long, strictfp, volatile, const, float, native, super, while},
    sensitive=true,
    morecomment=[l]{//},
    morecomment=[s]{/*}{*/},
    morestring=[b]",
    morestring=[b]',
    literate={«}{{\font\larm=larm1000\larm\char 190}}1
  }

%% force more space between subsections!
\makeatletter
\renewcommand\subsection{\medskip\@startsection{subsection}{2}{\z@}%
  {-.25ex\@plus -.1ex \@minus -.2ex}%
  {1.5ex \@plus .2ex \@minus -.4ex}%
  {\ifnum \scr@compatibility>\@nameuse{scr@v@2.96}\relax
    \setlength{\parfillskip}{\z@ plus 1fil}\fi
    \raggedsection\normalfont\sectfont\nobreak\size@subsection
  }%
}
\makeatother

%\lstdefinelanguage{xtext}{keywords={terminal, grammar, import, generate, hidden, as}, sensitive=true,string={"},morestring={'},comment={//},morecomment=[s]{/*}{*/}}

\lstset{tabsize=4, basicstyle=\sffamily\small, keywordstyle=\bfseries, columns=[r]fullflexible, escapechar={ß}}

%%\usepackage[english]{babel}

% import this as last package
\usepackage{hyperref}

\newlength{\itemindentlen}
\author{Sven Efftinge, et al.}
\title{Xtend2 Language Specification}


\begin{document}
\maketitle
\tableofcontents
\chapter{Preface}
\label{Preface}
This document specifies the language Xtend2. Xtend2 is a programming language implemented in Xtext, based on Xbase, and tightly integrated with Java.
It's main purpose is to write compilers, interpreters and other things, where you need to traverse typed tree structures (read EMF models). It
integrates and compiles to Java.  

Conceptually and syntactically, Xtend2 is a subset of Java, with the following differences:

\setlength{\itemindentlen}{\textwidth}
\begin{itemize}
\addtolength{\itemindentlen}{-2.5em}

\item \begin{minipage}[t]{\itemindentlen}
First class support for template syntax
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
extension methods
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
multiple dispatch aka polymorphic method invocation
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
built-in support for dependency injection
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
inferred return types
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
convenient graph navigation expressions
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
Uses Xbase expressions, which means :
		
\setlength{\itemindentlen}{\textwidth}
\begin{itemize}
\addtolength{\itemindentlen}{-2.5em}

\item \begin{minipage}[t]{\itemindentlen}
No checked exceptions
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
Pure object-oriented, i.e. no built-in types and no arrays
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
Everything is an expression, there are no statements
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
Closures
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
Type inference
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
Properties
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
Simple operator overloading
\end{minipage}
\item \begin{minipage}[t]{\itemindentlen}
Powerful switch expressions
\end{minipage}
\end{itemize}
\addtolength{\itemindentlen}{2.5em}
	
\end{minipage}
\end{itemize}
\addtolength{\itemindentlen}{2.5em}
\chapter{Language Concepts}
\label{LanguageConcepts}
On a first glance an Xtend2 file pretty much looks like a Java file. it starts with a package declaration followed by an import section, and after that comes the class definition.
That class in fact is directly translated to a Java class in the corresponding Java package.
Here is an example:
\begin{lstlisting}
package com.acme;

import java.util.List;

class MyClass {
	String first(List<String> elements) {
		...
	}
}

\end{lstlisting}

\section{Package Declaration}
\label{packageDecl}
Package declarations are like in Java, with the small difference, that an identifier can be escaped with a \textit{\textasciicircum{}} in case it conflicts with a keyword.

\subsection{Syntax}

\begin{lstlisting}
PackageDeclaration : 'package' QualifiedName;
QualifiedName : ID ('.' ID);

\end{lstlisting}




\section{Imports}
\label{imports}
The ordinary imports of type names are equivalent to the imports known from Java. Again one can escape any names conflicting with keywords using the \textit{\textasciicircum{}}.

\subsection{Importing Extension Methods}
\label{importExtension}
In addition one can add the keyword \lstinline]°extension° after the import (where the keyword \lstinline]°static° occurs in Java), which makes any static methods 
from a certain type available as an 
\hyperref[extensionMethods]{extension method~(§\ref*{extensionMethods})}. The \lstinline]°extension° keyword only works with qualified imports (i.e. non-wildcard imports).
If you only want to make some methods from the imported type available using the extension syntax, you can provide a list of function names.
You cannot further specify type arguments.

Example:

\begin{lstlisting}
import extension java.util.Collections.(sort, shuffle);

\end{lstlisting}


This will import four methods (\lstinline]°sort(List<T>)°, \lstinline]°sort(List<T>, Comparator<? super T>)°, \lstinline]°shuffle(List<?>)°, \lstinline]°shuffle(List<?>, Random)°) available as 
\hyperref[extensionMethods]{extension methods~(§\ref*{extensionMethods})}.


\subsection{Syntax}

\begin{lstlisting}
ImportSection : (Import|Injection)*;
Import : 'import' 'extension'? QualifiedName (('.''*')|'('ID (',' ID)')')?;

\end{lstlisting}




\section{Dependency Injection}
\label{inject}
Xtend2 has built-in support for dependency injection, which means that declaring dependencies is a first class feature of Xtend2.
It is done as part of the import section and looks like this:

\begin{lstlisting}
inject MyService myService;

\end{lstlisting}


This will translate to Java as a field as well a setter method annotated with @Inject: 

\begin{lstlisting}[language=Java]
private MyService myService;

@Inject
public void setMyService(MyService myService) {
	this.myService = myService;
}

\end{lstlisting}


\subsection{Name can be inferred}

You could even leave the name out as by default it will be the lower case version of the simple name of the class.
That is the following inject declaration would be equvalent with the former.

\begin{lstlisting}
inject MyService;

\end{lstlisting}



\subsection{Guice Keys}

Xtend2 uses Guice as the dependency injection container, which in turn uses types as keys. Those types can be generified, which
is why you can write the following in Xtend2.

\begin{lstlisting}
inject HashMap<String,MyType> myCache;

\end{lstlisting}


Also Guice allows to use annotations to further distinct between different implementations of the same type. This is usually done by using
the \lstinline]°com.google.inject.name.Named° annotation or a custom annotation which is itself annotated with \lstinline]°com.google.inject.BindingAnnotation°.

Xtend supports these two variants:

\begin{lstlisting}
inject @CustomAnnotation my.Service customService; 
inject @"NamedThingy" my.Service namedService;

\end{lstlisting}


Providers are declared like in plain Guice. That is you wrap your type into \lstinline]°com.google.inject.Provider°.
Example:

\begin{lstlisting}
inject @CustomAnnotation Provider<Service> customServiceProvider; 

\end{lstlisting}



\subsection{Injected Extension Methods}

The \lstinline]°extension° keyword is also available for \lstinline]°inject° which is much nicer than importing static methods, because you don't bind your code to 
actual implementation, since the implementation can be easily changed through the global dependency injection configuration (i.e. Guice modules).

Example:

Imagine you had something like IterableExtensions which adds some higher-order functions
to the java.util collection library (we will likely ship a library based on google guava).

\begin{lstlisting}
class IterableExtensions {
   public <T> T find(Iterable<T> elements, Function<T,Boolean> func) {
     ...
   }
}

\end{lstlisting}


You declare the dependency like this:

\begin{lstlisting}
inject extension my.common.IterableExtensions;

\end{lstlisting}
  

and can then use it as if those functions were members of any \lstinline]°java.lang.Iterable°:

\begin{lstlisting}
myIterable.find(e|e.name=="Foo")

\end{lstlisting}


The nice thing with using injection as opposed to static methods is, that in case there is a bug in the extension or it is implemented inefficiently or you 
just need a different strategy, you can simply exchange the component with another implementation. You do this without modifying the library nor the client code. 
You'll only have to change the binding in your guice module. Also this gives you a general hook for any AOP-like thing you would want to do, 
or allows you to write against an SPI, where the concrete implementation can be provided by a third party.

\begin{lstlisting}
ImportSection : (Import|Injection)*;
Import : ...
Injection : 'inject' 'extension'? ('@'STRING | '@'TypeReference) TypeReference ('('ID (',' ID)')')?;

\end{lstlisting}




\section{Class Declaration}
\label{classDeclaration}
The class declaration reuses a lot of Java's syntax but still is a bit different in some aspects.
Firstly the default visibility of any class is \lstinline]°public°. It is possible to write it explicitly but if not specified it defaults to public.
Java's default "package private" visibility does not exist.

The \lstinline]°abstract° as well as the \lstinline]°final° modifiers are directly translated to Java, hence have the exact same meaning.

\subsection{Inheritance}

Also inheritance is directly reused from Java. Xtend2 allows single inheritance of Java classes as well as implementing multiple Java interfaces.


\subsection{Generics}

Full Java Generics with the exact same syntax and semantics are supported.


\subsection{Abstract Classes}

A class is automatically \lstinline]°abstract° if one of its functions is abstract. 
This is the case if the function's expression (body) is not defined. 


\subsection{Syntax}

\begin{lstlisting}
ClassDeclaration : 
	(Visibility|'abstract'|'final')* 'class' TypeParameters? ('extends' QualifiedName)? ('implements' QualifiedName (',' QualifiedName)*)? '{'
		Members*
	'}'
;
Visibility : 'public'|'protected'|'private';


\end{lstlisting}




\section{Functions}
\label{Functions}
Xtend2 functions are declared within a class and are usually translated to a corresponding Java method with the exact same signature. 
The only exception is overloaded methods, which compile to a single method. This is explained in the section about 
\autoref{polymorphicDispatch}. 

An example of a function declaration
\begin{lstlisting}
Boolean equalsIgnoreCase(String s1,String s2) :
	s1.toLowerCase() == s2.toLowerCase();

\end{lstlisting}


\subsection{Visibility}

The default visibility of a function is \lstinline]°public°, which can also be declared explicitly. The two other available visibilities are \lstinline]°protected° and \lstinline]°private°.


\subsection{Overriding Functions}

Like Java methods an Xtend2 function can be declared non overridable using the keyword \lstinline]°final°.

Also if a function overrides a method from the suberclass, the \lstinline]°override° keyword is mandatory.

Example:
\begin{lstlisting}
final override Boolean equalsIgnoreCase(String s1,String s2) :
	s1.toLowerCase() == s2.toLowerCase();

\end{lstlisting}



\subsection{Abstract Functions}

A function is automatically considered \lstinline]°abstract° if its expression (body) is not defined.

Example:
\begin{lstlisting}
Boolean myAbstractFunction(String s1,String s2);

\end{lstlisting}



\subsection{Inferred Return Types}
\label{inferredReturnTypes}
If the return type of a function can be inferred it does not need to be declared.
That is the function

\begin{lstlisting}
Boolean equalsIgnoreCase(String s1,String s2) :
	s1.toLowerCase() == s2.toLowerCase();

\end{lstlisting}


could be declared like this:

\begin{lstlisting}
equalsIgnoreCase(String s1,String s2) :
	s1.toLowerCase() == s2.toLowerCase();

\end{lstlisting}


This doesn't work for abstract function declarations as well as if the 
return type of a function depends on a recursive call of the same function.
The compiler tells the user when it needs to be specified.


\subsection{Generics}

Full Java Generics with the exact same syntax and semantics are supported.


\subsection{Polymorphic Method Overloading}
\label{polymorphicDispatch}
It is possible to overload methods, but overloaded methods are not invoked based on the static types of arguments as in Java.
In Xtend2 an overloaded method is invoked based on the rutime types of the arguments. This has a lot of advantages and is a very convenient way 
to add functionality to heterogenous data structors especially with deeper type hierarchies (i.e. EMF models and other typed domain models).
It essentially eleminates the problem the visitor pattern is trying to solve.

This allows you to write a code generator by defining a function for all kinds of AST elements you want to process.

\begin{lstlisting}
class MyCompiler {
	compile(CompilationUnit cu) :
		... cu.types.compile() ...;
	compile(Interface interface) :
		... cu.members.compile() ...;
	compile(Class cu) :
		... cu.members.compile() ...;
	compile(Field cu) :
		...;
	...
}

\end{lstlisting}


\subsubsection{How overloaded method are translated to Java}

The polymorphic behavior of overloaded methods shall be transparently work no matter you call such a function from Java, Xtend2 or any other JVM language.
Therefore the runtime dispatch is not done on the caller's side but on the declaration side. 
For each set of methods where the name is equal and the number of arguments is equal, the most common dominator signature is taken 
or computed if there is no most common denominator. Only for that signature a Java method is generated. 
Within the implementation the correct method is looked up at runtime using if else cascades.
This is done by sorting the methods from most specific to least specific and generating an if-else cascade for the code.

Example: The following functions
\begin{lstlisting}
foo(Number x) : 'it's some number';
foo(Integer x) : 'it's an int';

\end{lstlisting}
 

compile to the following Java method:

\begin{lstlisting}[language=Java]
public String foo(Number x) {
	if (x== null || x instanceof Integer) {
		return "it's an int";
	} else {
		return "it's some number";
	}
} 

\end{lstlisting}


In case there is no single most general signature, one is computed and the
different overloaded methods are matched in the order they are declared within the class file.
Example:

\begin{lstlisting}
foo(Number x, Integer y) : 'it's some number and an int';
foo(Integer x, Number x) : 'it's an int and a number';

\end{lstlisting}


\begin{lstlisting}[language=Java]
public String foo(Number x) {
	if ((x== null || x instanceof Number) && (y== null ||y instanceof Integer)) {
		return "it's some number and an int";
	} else if ((x== null || x instanceof Integer) && (y== null || y instanceof Number)){
		return "it's an int and a number";
	} else {
		throw new UnsupportedOperationException("foo is not implemented for arguments "+x+" and "+y);
	}
} 

\end{lstlisting}


As you can see a null reference is always a match. If you want to fetch null you can declare a parameter using the type \lstinline]°java.lang.Void°.

\begin{lstlisting}
foo(Number x) : 'it's some number';
foo(Integer x) : 'it's an int';
foo(Void x) : throw new NullPointerException("x");

\end{lstlisting}


Which compiles to the following Java code:

\begin{lstlisting}[language=Java]
public String foo(Number x) {
	if (x== null) {
		throw new NullPointerException("x")
	} else if (x== null || x instanceof Integer) {
		return "it's an int";
	} else {
		return "it's some number";
	}
} 

\end{lstlisting}


Note that the type \lstinline]°java.lang.Void° is considered the most specific type and will always have higher precedence than any other types in the ordering.


\subsubsection{Overloading Functions from Super Types}

Functions (or Java Methods) from super types are not included in the dispatch. You'll have to call them explicitly using the \lstinline]°super° reference. 


\subsubsection{Excluding functions from polymorphic dispatch}

Sometimes you need to have the static dispatch from Java, for instance if you implement an interface with two method declarations where one overloads 
the other. In that case you have to add the keyword \lstinline]°no-dispatch°.


\subsection{Syntax}

Syntactically Xtend functions are much like Java methods, expect that there are no static methods, the return types are optional and of course the function body
consists of one expression instead of a sequence of statements. 

\begin{lstlisting}
FunctionDef : ('public'|'protected'|'private'|'abstract'|'final')* TypeParameters? TypeRef? ID'('(ParameterDeclaration (',' ParameterDeclaration)*)?')' 
	('throws' TypeRef (',' TypeRef)*)?
    ( ':' Expression ';'
	| BlockExpression
	| ';')

\end{lstlisting}




\section{Expressions (in addition Xbase)}
\label{Expressions}
Xtend2 adds a couple of expressions to the basic set of expressions provided by Xbase.

\subsection{Rich Strings}
\label{RichStrings}
Of course there is the template expression, which is used to write readable string concatenation, which is the main thing you do when writing a code generator.
Xtend2 reuses the syntax known from the well known and widely used Xpand template language (in fact Xtend2 is considered the successor to Xpand and Xtend). Let's have a look at
an example of how a typical function with template expressions look like:

\begin{lstlisting}
toClass(Entity this) :»
	package «packageName»;
	
	«placeImports»
	
	public class «name» «IF extendedType!=null»extends «extendedType»«ENDIF»{
		«FOREACH members»
			«member.toMember»
		«ENDFOREACH»
	}
	«;

\end{lstlisting}


If you are familiar with Xpand, you'll notice that it is exactly the same syntax. The difference is, that the template syntax is actually an expression, which means it 
can occur everywhere where an expression is expected. For instance in conjunction the powerful switch expression from Xbase:

\begin{lstlisting}
toMember(Member this) :
	switch(this) {
		Field :»private «type» «name» ;«;
		Method case isAbstract :» abstract «...;
		Method:» ..... «;
	};

\end{lstlisting}


\subsubsection{IF in Rich Strings}
\label{RichStringIF}
There is a special IF to be used within rich strings which is identical in syntax and meaning to the old IF from Xpand.
Note that you could also use the if expression, but since it has not an explicit terminal token, it is not as readable in that context.


\subsubsection{FOREACH in Rich Strings}
\label{RichStringFOREACH}
Also the FOREACH statement is available and can only be used in the context of a rich string.
It also supports the SEPARATOR and ITERATOR declaration from Xpand. 
\todo[inline]{Think about whether we really want to stick to the verbose syntax form previous Xpand for SEPARATOR and ITERATOR}


\subsubsection{Typing}
\label{RichStringType}
The rich string is translated to an efficient string concatenation and the return type of a rich string is \lstinline]°java.lang.CharStream° which allows
much room for efficient implementation.


\subsection{Extension Method Syntax}
\label{extensionMethods}
Static functions as well as any members annotated with @Extension can be called using the extension method syntax.
This means that a function imported through a static import, like e.g. java.util.Collections.singleton(T) can be invoked using 
the member syntax.

Example:

\begin{lstlisting}
"Foo".singleton

\end{lstlisting}


is the same as

\begin{lstlisting}
singleton("Foo")

\end{lstlisting}


Note that extension methods never shadow a member of the current reference. That is if \lstinline]°java.lang.String° has a field \lstinline]°singleton°, a method \lstinline]°singleton()° or a method
\lstinline]°getSingleton()°, those memebers would be referenced. That is done at compile time, so the tooling is able to tell you what you actually reference.

Static functions as well as extensions in the previous version of Xtend make clients not only depend on a certain signature but on the implementation as it is not 
possible to exchange the implementation of a static function. That's where the @Extension annotation in conjunction with dependency injection comes in.

If you have declared a field with @Extension than all members of that type become available for extension method syntax.

Example

Imagine the following Java interface

\begin{lstlisting}
interface MyExtensions {
	String getComputedProperty(SomeType type);
}

\end{lstlisting}


With Java you would have to call this method on an instance of MyExtensions like so:

\begin{lstlisting}
myExtensions.getComputedProperty(someType)

\end{lstlisting}


In Xtend2 you can have the instance injected like in Java:

\begin{lstlisting}
@Inject @Extension MyExtensions extensions;

\end{lstlisting}


But instead of using the long expression known from Java (which would also work) you are able to use the extension method syntax:

\begin{lstlisting}
someType.computedProperty

\end{lstlisting}


It will statically bind to the right method and even more important it doesn't bind to a specific implementation but just to the signature.
The implementation can be provided through dependency injection.


\chapter*{List of External Links}


\listoftodos
\end{document}
